---
title: "PlotsForPoster"
output: html_document
date: "2025-02-18"
---

# Set up
```{r}
rm(list=ls())
library(ggplot2)
library(tidyr)
library(tidyverse)
library(WGCNA)
library(pheatmap)
library(clusterProfiler)
library(org.Hs.eg.db)
source("~/Documents/Projects/clarava_manuscript_2025/WGNCA/WGNCA_Helper.R")
```

# Loading in TPM and Meta
```{r}
AllMeta<-read.csv("/Users/alinaklineschoder/Desktop/Data/Clarava_Meta.csv")%>%
  column_to_rownames(var="Sample_ID")
AlliSort<-read.csv("/Users/alinaklineschoder/Desktop/Data/Clarava_iSort.csv")%>%
  column_to_rownames(var="Sample_ID")
AllTPM<-read.csv("/Users/alinaklineschoder/Desktop/Data/Clarava_TPM.csv")%>%
  column_to_rownames(var="Gene")%>%
  dplyr::select(-c(X))
```

# All Data

# Pre-process TPM
```{r}
tpm_data <- t(AllTPM)
tpm_data <- tpm_data[,apply(tpm_data, 2, var) > 0.1]
log_tpm_data <- log2(tpm_data + 1)
gsg = goodSamplesGenes(log_tpm_data, verbose = 3)
log_tpm_data <- log_tpm_data[gsg$goodSamples,gsg$goodGenes]
```

# detect sample outliers
```{r}
sample_tree <- hclust(dist((log_tpm_data)), method = "average")
plot(sample_tree, main = "Sample Clustering to Detect Outliers", sub = "", xlab = "", cex = 0.5)
```

```{r}
powers = c(1:20)
sft = pickSoftThreshold(log_tpm_data, powerVector = powers, verbose = 5)

par(mfrow = c(1,2))
plot(sft$fitIndices[,1], -sign(sft$fitIndices[,3]) * sft$fitIndices[,2],
     xlab = "Soft Threshold (power)", ylab = "Scale-Free Topology Fit", type = "o",
     main = "Scale Independence")
abline(h = 0.9, col = "red")

plot(sft$fitIndices[,1], sft$fitIndices[,5], type = "o",
     xlab = "Soft Threshold (power)", ylab = "Mean Connectivity",
     main = "Mean Connectivity")
```

```{r}
softPower = 6
adjacency = adjacency(log_tpm_data, power = softPower)
TOM = TOMsimilarity(adjacency)
dissTOM = 1 - TOM
gene_tree = hclust(as.dist(dissTOM), method = "average")
plot(gene_tree, main = "Clustering of Genes", xlab = "", sub = "")
```

```{r}
dynamicMods = cutreeDynamic(dendro = gene_tree, distM = dissTOM,
                            deepSplit = 2, pamRespectsDendro = FALSE,
                            minClusterSize = 30)
dynamicColors = labels2colors(dynamicMods)
plotDendroAndColors(gene_tree, dynamicColors, "Dynamic Tree Cut",
                    main = "Gene Module Identification", dendroLabels = FALSE)
```

```{r}
ME_list = moduleEigengenes(log_tpm_data, colors = dynamicColors)
MEs = ME_list$eigengenes
METree = hclust(dist(MEs), method = "average")
plot(METree, main = "Clustering of Module Eigengenes", xlab = "", sub = "")
```

# Comparing with iSort - granular
```{r}
broad_cells_of_interest<-c("Neutrophils","Monocytes","Dendritic_cells_activated","T_cells_gamma_delta")

isort_results_val<-AlliSort[,broad_cells_of_interest]
common_samples <- intersect(rownames(MEs), rownames(isort_results_val))
MEs <- MEs[common_samples,]
isort_results_val <- isort_results_val[common_samples,]

module_trait_cor = cor(MEs, isort_results_val, use = "p")
module_trait_pval = corPvalueStudent(module_trait_cor, nSamples = nrow(MEs))

pheatmap(module_trait_cor, display_numbers = round(module_trait_cor, 3), cluster_rows = TRUE, cluster_cols = TRUE)
```

# Identifying Hub Genes
```{r}
# Calculate module membership (MM) and gene significance (GS)
MM = cor(log_tpm_data, MEs, use = "p")
GS = cor(log_tpm_data, isort_results_val, use = "p")

hub_gene_threshold = 0.7
hub_genes_per_module <- list()
names(dynamicColors) <- colnames(log_tpm_data)
for (module in colnames(MEs)) {
  module_genes <- names(which(dynamicColors == gsub("ME","",module)))
  module_MM <- MM[module_genes, module]
  module_GS <- GS[module_genes, , drop = FALSE]
  hub_genes <- names(which(module_MM > hub_gene_threshold & apply(module_GS, 1, max) > hub_gene_threshold))
  hub_genes_per_module[[module]] <- hub_genes
}

# Save hub genes per module
hub_genes_df <- do.call(rbind, lapply(names(hub_genes_per_module), function(module) {
  data.frame(Module = module, Gene = hub_genes_per_module[[module]])
}))

print(paste("Identified hub genes for", length(hub_genes_per_module), "modules"))
